Testing Framework and Reference

### 2025-05-13 nbs/development/testing.qmd
docs(development): update parameter recovery validation requirements
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

| AMDiRE Element ID | Element Type | Element Description | Test IDs | Coverage Status |
|-------------------|--------------|---------------------|----------|----------------|
| RG-01 | Research Goal | Probabilistic RNA Velocity | TEST-BDD-01, TEST-UNIT-05 | Complete |
| RG-02 | Research Goal | Uncertainty Quantification | TEST-BDD-02 | Partial |
| FN-03 | Functional Requirement | Model Training | TEST-BDD-04 | Complete |
| FN-05 | Functional Requirement | Velocity Computation | TEST-BDD-05 | Complete |
| CMP-01 | Component | PyroVelocityModel | TEST-BDD-06 | Complete |
| CMP-02 | Component | DynamicsModel | TEST-UNIT-05, TEST-BDD-01 | Complete |
| MM-01 | Mathematical Model | RNA Velocity Dynamics | TEST-UNIT-08 | Complete |

@@

| Module | Feature | Status | Description | Related AMDiRE Elements |
|--------|---------|--------|-------------|------------------------|
| dynamics_model | Standard dynamics model computes expected counts | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes expected unspliced and spliced counts | MM-01, CMP-02, FN-05 |
| dynamics_model | Standard dynamics model computes steady state | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes steady state values | MM-01, CMP-02 |
| dynamics_model | Legacy dynamics model matches legacy implementation | ✅ Implemented | Verify that the LegacyDynamicsModel matches the legacy implementation | CMP-02, TC-04 |
| dynamics_model | Dynamics model handles edge cases | ✅ Implemented | Verify that the dynamics model handles edge cases gracefully | CMP-02, REL-02 |
| dynamics_model | Dynamics model with library size correction | ✅ Implemented | Verify that the dynamics model correctly applies library size correction | CMP-02, SC-01 |
| prior_model | LogNormal prior model creates stochastic nodes | ✅ Implemented | Verify that the LogNormalPriorModel creates appropriate stochastic nodes | CMP-03, ALG-02 |
| likelihood_model | Poisson likelihood model defines distributions | ✅ Implemented | Verify that the PoissonLikelihoodModel defines appropriate Poisson distributions | CMP-04, ALG-02 |
| model | PyroVelocity model integrates components | ✅ Implemented | Verify that the PyroVelocityModel correctly integrates all components | CMP-01, SG-01 |
| model | PyroVelocity model trains on AnnData | ✅ Implemented | Verify that the PyroVelocityModel can train on AnnData objects | CMP-01, FN-03, INT-01 |

@@

PyroVelocity includes a validation framework (PRD-678-VALID) to verify that the modular model can replicate the legacy model. This framework is a critical part of our testing strategy, ensuring that the refactored implementation maintains compatibility with existing workflows.

@@

- **Direct Comparison**: Comparing outputs between legacy and modular implementations
- **Synthetic Data Validation**: Using synthetic data with known properties
- **Real Data Validation**: Using downsampled real datasets (e.g., `tests/data/preprocessed_pancreas_50_7.json`)
- **Parameter Validation**: Ensuring parameter inference matches between implementations
- **Shape Validation**: Verifying tensor shapes are consistent between implementations

@@

- Direct comparison tests for model outputs
- Metrics for quantifying differences between implementations
- Visualization tools for analyzing differences
- Utilities for creating patched versions of components to address compatibility issues

@@

- **TC-04**: Backward Compatibility constraint
- **REQ-01**: Parameter Estimation Accuracy requirement
- **SG-03**: Comprehensive Testing system goal
- **REL-01**: Training Convergence reliability requirement

>>>>

| AMDiRE Element ID | Element Type | Element Description | Test IDs | Coverage Status |
|-------------------|--------------|---------------------|----------|----------------|
| RG-01 | Research Goal | Probabilistic RNA Velocity | TEST-BDD-01, TEST-UNIT-05 | Complete |
| RG-02 | Research Goal | Uncertainty Quantification | TEST-BDD-02 | Partial |
| RG-03 | Research Goal | Parameter Inference | TEST-UNIT-01, TEST-BDD-07 | Planned |
| FN-03 | Functional Requirement | Model Training | TEST-BDD-04 | Complete |
| FN-05 | Functional Requirement | Velocity Computation | TEST-BDD-05 | Complete |
| CAP-06 | Capability | Parameter Recovery Validation | TEST-BDD-07, TEST-UNIT-09 | Planned |
| CAP-07 | Capability | Model Comparison | TEST-BDD-08, TEST-UNIT-10 | Planned |
| CMP-01 | Component | PyroVelocityModel | TEST-BDD-06 | Complete |
| CMP-02 | Component | DynamicsModel | TEST-UNIT-05, TEST-BDD-01 | Complete |
| SC-03 | Scientific Constraint | Parameter Identifiability | TEST-BDD-07 | Planned |
| MM-01 | Mathematical Model | RNA Velocity Dynamics | TEST-UNIT-08 | Complete |
| ALG-02 | Algorithm | Parameter Recovery Validation | TEST-UNIT-09 | Planned |

@@

| Module | Feature | Status | Description | Related AMDiRE Elements |
|--------|---------|--------|-------------|------------------------|
| dynamics_model | Standard dynamics model computes expected counts | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes expected unspliced and spliced counts | MM-01, CMP-02, FN-05 |
| dynamics_model | Standard dynamics model computes steady state | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes steady state values | MM-01, CMP-02 |
| dynamics_model | Legacy dynamics model matches legacy implementation | ✅ Implemented | Verify that the LegacyDynamicsModel matches the legacy implementation | CMP-02, TC-04 |
| dynamics_model | Dynamics model handles edge cases | ✅ Implemented | Verify that the dynamics model handles edge cases gracefully | CMP-02, REL-02 |
| dynamics_model | Dynamics model with library size correction | ✅ Implemented | Verify that the dynamics model correctly applies library size correction | CMP-02, SC-01 |
| prior_model | LogNormal prior model creates stochastic nodes | ✅ Implemented | Verify that the LogNormalPriorModel creates appropriate stochastic nodes | CMP-03, ALG-02 |
| likelihood_model | Poisson likelihood model defines distributions | ✅ Implemented | Verify that the PoissonLikelihoodModel defines appropriate Poisson distributions | CMP-04, ALG-02 |
| model | PyroVelocity model integrates components | ✅ Implemented | Verify that the PyroVelocityModel correctly integrates all components | CMP-01, SG-01 |
| model | PyroVelocity model trains on AnnData | ✅ Implemented | Verify that the PyroVelocityModel can train on AnnData objects | CMP-01, FN-03, INT-01 |
| validation | Parameter recovery validation | 🔄 Planned | Verify that the model can recover known parameters from synthetic data | CAP-06, RG-03, SC-03, ALG-02 |
| validation | Model comparison | 🔄 Planned | Verify that different model implementations can be compared | CAP-07, TC-04, SG-03 |

@@

PyroVelocity includes a validation framework (PRD-678-VALID) to verify that the modular model can replicate the legacy model and to validate the model's ability to recover known parameters from synthetic data. This framework is a critical part of our testing strategy, ensuring that the refactored implementation maintains compatibility with existing workflows and that the model is internally consistent.

@@

- **Direct Comparison**: Comparing outputs between legacy and modular implementations
- **Synthetic Data Validation**: Using synthetic data with known properties
- **Real Data Validation**: Using downsampled real datasets (e.g., `tests/data/preprocessed_pancreas_50_7.json`)
- **Parameter Validation**: Ensuring parameter inference matches between implementations
- **Shape Validation**: Verifying tensor shapes are consistent between implementations
- **Parameter Recovery Validation**: Validating model's ability to recover known parameters from synthetic data

@@

- Direct comparison tests for model outputs
- Metrics for quantifying differences between implementations
- Visualization tools for analyzing differences
- Utilities for creating patched versions of components to address compatibility issues
- Parameter recovery validation tools for generating synthetic data and evaluating parameter recovery

## Parameter Recovery Validation

Parameter recovery validation is a fundamental technique in probabilistic modeling that assesses whether a model's inference procedure can correctly recover known parameter values from synthetic data generated by the model itself. It serves as a crucial sanity check before applying models to real-world data where true parameter values are unknown.

The parameter recovery validation workflow follows these steps:

1. **Parameter Generation**: Sample parameter sets from prior distributions
2. **Synthetic Data Generation**: Create synthetic data using known parameters
3. **Inference**: Perform inference on synthetic data to estimate parameters
4. **Evaluation**: Compare estimated parameters to original known values
5. **Metrics Computation**: Calculate recovery accuracy, precision, and coverage metrics

The validation framework provides tools for each step of this workflow, including:

- Functions for generating parameter sets from prior distributions
- Functions for creating synthetic data with known parameters
- Functions for performing inference on synthetic data
- Functions for comparing estimated parameters to true values
- Functions for calculating and visualizing recovery metrics

@@

- **TC-04**: Backward Compatibility constraint
- **REQ-01**: Parameter Estimation Accuracy requirement
- **SG-03**: Comprehensive Testing system goal
- **REL-01**: Training Convergence reliability requirement
- **CAP-06**: Parameter Recovery Validation capability
- **CAP-07**: Model Comparison capability
- **RG-03**: Parameter Inference research goal
- **SC-03**: Parameter Identifiability scientific constraint
